Mining Interaction Log Data in a Creative Arts Practice using Macro Gestures and Transition Matrices

### 2015-09-21 flux-entropy-paper/flux-entropy-paper.tex
intro updates

\begin{abstract}

%consider the following? .............

%   %Digital arts practices can allow for the archival of interaction
% log data. Such log data can capture and preserve performances of
% new-media artists and musicians and %and can facilitate post-hoc data
% processing and data mining.

  In this note we describe the analysis of a data archive for an iPad
  musical-ensemble performance practice where transition matrix
  measures are able to distinguish between sessions of different
  contexts, and the internal structure of these sessions. The archive
  includes 95 touch data logs of ensemble touch-screen interactions in
  concerts, rehearslas and formal HCI studies of groups of musicians
  over more than two years.

  % At the lowest level, data for each musician in a particular
  % performance was archived as Apple iOS callback methods for
  % touching and swiping.

  The data analysis has been undertaken by first classifying
  these data into higher level gesture types and then describing the
  transitions between these gestures for each artist as a function of
  time. We find that a transition-matrix measure, which we call
  ``flux'', can be used to distinguish different performance types and
  contexts, as well as different phases (beginning, middle and end) of
  all of the performances in our corpus.

\end{abstract}

@@

%\section{General comments}

% \begin{itemize}
% \item we should be clear (and careful) where we're talking about
%   groups TMs, and where we're talking about individual performers
% \item what's the terminology we're using for a ``performance''? Is it
%   ``performance'', or is that too music-specific? ``Session''?
%   Alternatives are ``activity'', or ``interaction'', but they're so
%   general\ldots This decision has implications for our other
%   terminology: participant vs performer vs musician, etc.
% \item Title: maybe ``Traces through time: making sense of longitudinal
%   open-ended interaction data'' or some such? I think that's our big
%   contribution: it takes the in-the-wild, practitioner-driven research
%   program stuff and tries to make some quantitative sense out of it
%   post-hoc.
% \end{itemize}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figures/three-performance-contexts}
  \caption{Collaborative musical interaction with our touch-screen
    interfaces in three different contexts: (left to right) rehearsal
    in a studio,
    performance in front of an audience, and demonstration at a
    research poster session. All of these session were improvised,
    and in the performance session, the musicians used percussion
    instruments as well as iPads.
    \label{fig:three-performance-contexts}}
\end{figure*}

Modern computing devices and interfaces continue to open up open-ended
creative opportunities for new artistic experiences. Understanding and
evaluating interfaces in these contexts presents many
challenges~\cite{Resnick:2005yu,Shneiderman:2007qv}.

In the context of live performance, interface evaluation, using, for
example, post-performance questionnaires, can omit consideration of
the dynamic needs of artists in the moment of the performance and in
the wild of the performance context.

In contrast to traditional art forms, logged interaction data from
interaction with a computer-based artistic interface can provide an
enduring record of that performance for post-hoc quantitative
investigation as well as for curation and
preservation~\cite{England:2014ys}.

Over time, the accumulation of
interaction log data might allow data-mining techniques to be deployed
to potentially describe universal features of such interaction.

It is a feature of creative expression in the arts that highly skilled
practitioners exploit simple tools to produce works of beauty whether
it be static paintings or dynamic performance.

In previous work~\cite{Martin:2014cr} researchers have shown that
trained percussionists can exploit a very simple iPad interface to
discover new categories of musical gestures.

In that paper, iPad apps
were studied where tapping was mapped to produce produced short
sounds, while swiping produced long sounds with the volume
proportional to the velocity of the moving touch point.

It turned out that trained percussionists working together in an
ensemble were able to discover and repeatedly make use of five macro
gestures made up of taps and swipes: fast taps, slow taps, random
taps, short swipes and swirls. This process of discovery of
interesting musical gestures from a simple interface is analogous to
what percussionists do with acoustic instruments -- drums and bowls
and bells are often tapped, stroked and hit in novel ways to
communicate meaning and emotion through music~\cite{Schick:2006fk}.

In this note we describe a log data archive that has been accumulated
to support a long-term practice-led research program into simple iPad
interfaces for group music-making similar to those described in Martin
et al~\cite{Martin:2014cr}. Over a period of more than two years
(April 2013--June 2015), we collected data from group music-making
sessions where the participants used custom-built, touch-screen apps
on Apple iPad devices. 
Although several different apps have been
included in our corpus, they all shared the simple mode of interaction
where the majority of the screen could be used for free-form touch
interactions; tapping produced short sounds, while swiping or swirling
produced long sounds with the volume proportional to the velocity of
the moving touch point. Apart from this commonality, our apps featured
different palettes of available sounds and modes of synthesis,
different arrangements of sounds on the iPad screens, different kinds
of visual display and feedback, different kinds of networked
interactions (between the iPads in an ensemble and with a central
server), as well as a variety of special features for each app.

In all of our data-collection sessions, every touch
interaction---every touch down, dragging movement and touch up---was
time-stamped and logged to disk for later analysis. These sessions
included rehearsals without an audience as well as live performances
at art festivals and other events. They also included two
formally-structured HCI experiments in order to understand and iterate
on the system and interface design. Some of the performances were
iPad-only and others used iPads together with other acoustic
percussion instruments. Although the majority of the sessions were
free-form improvisations, a number of rehearsals and performances
involved composed pieces where the musicians read and interpreted
percussion notation as they used the apps. Representative photographs
of some of these session contexts (rehearsal, performance including
other acoustic instruments and demonstrations) are are shown
Figure~\ref{fig:three-performance-contexts}) and other descriptive
statistics about the sessions are shown in Table \ref{corpus-table}.
In our corpus, the median length of sessions is 7m29s and the median
number of participants per ensemble is four; . Metadata is included
together with the interaction data for each session to record the
\emph{context} (performance, rehearsal, study), session \emph{type}
(improvisation or composition), and \emph{instrumentation}
(iPads-only, or iPads with acoustic percussion). The distribution of
recorded sessions by context and type is shown in Figure
\ref{fig:count-data}.

% TODO fix this table to remove descriptive data about the
% demonstration sessions.
\begin{table}
\centering
\begin{tabular}{l|llll}
\hline
            & Total & Min  & Median   & Max     \\ 
\hline
N           & 150       &         &        &  \\
%Record size & 306MB     &         &        &   \\
Length      & 20H34M22S & 1M0S    & 7M29S  & 26M0S\\
Participants& 524       & 1       & 4      & 9    \\
Flux        &           & 0.02239 & 0.1445 & 0.3246\\
%Entropy     &           &  &  & \\          
\hline
\end{tabular}
\caption{
  Representative data about the corpus of musical touch-screen
  interaction sessions used in this study. These sessions were
  rehearsals, performances, studies, and demonstration sessions and 
  participants improvised or performed compositions. Each record
  consisted of touch-data recorded in CSV format and was later analysed
  with a gestural classification system. Some participants were
  present in multiple sessions, however this data was not tracked in
  the corpus.\label{corpus-table}}
\end{table}

@@

Suppose that each participant's activity in a collaborative
interaction can at any given time be classified as belonging to one of
$m$ distinct \emph{states}. Over the course of the activity, each
participant $X$ generates a length $N$ sequence of these interaction
states
\begin{equation}
 X_n \hskip 2em n = 1, \ldots, N
\end{equation}
which can be thought of as Markov process. If there is an
(automated) classification process for extracting each participant's
state sequence from (or during) their interaction data, then we can
estimate the $m \times m$ transition matrix $P$ which characterises
this process by simply counting the number of times state $j$ follows
state $i$
\begin{equation}
  p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
\end{equation}
The transition activity of the whole ensemble can be calculated by
averaging the TM for each performer. The matrix can be normalised so
that the sum of all elements is equal to one.

@@

To do this we calculate two high-level quantities from each session:
``flux'' and ``entropy''. The flux of a sequence is defined as the
ratio of state transitions (e.g. $A \rightarrow B$ where $A \neq B$),
to self transitions (e.g. $A \rightarrow A $). The flux of a TM $P$ is
given by $\mathrm{flux}(P) = 1 - \mathrm{tr}(P)$ where the trace
$\mathrm{tr}(P)$ is the sum of the diagonal entries of $P$.
Intuitively, flux is a measure of how frequently the participant/group
changes state. Flux returns a value in the interval $[0,1]$ and will
return 0 when participants never change state, and 1 when participants
never stay in the same state for two measurements in a row.

Another useful measure that can be used on a transition matrix is its
entropy, defined in the information-theoretic\cite{Shannon:1948rt}
sense:
\begin{equation}
  H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})
\end{equation}
This entropy measure is small when the matrix is sparse, and largest
when each matrix element is equal. It offers a different perspective
on collaborative interactions than the flux measure by capturing the
breadth of the gestural space explored by the participants throughout
the course of an interaction. Consider the degenerate case of a
participant who alternates between two states for a whole performance:
the flux in this case will be maximal ($\mathrm{flux}(P) = 1$) since
there are no self-transitions, (only $A \rightarrow B$ or
$ B \rightarrow A$) even though the participant has only used a small
subset of the state space. The entropy of this interaction, on the
other hand, will be low. Entropy, therefore, is a measure of how broad
the participant's exploration of the state space is in a given
interaction.

@@

%  Secondly, do these performance measure allow us
% further insight into the corpus than the metadata can provide?
% % TODO I think this second question is too broad?

@@

The distribution of flux and entropy with respect to session contexts,
types and instrumentation are shown in Figure
\ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
performed on the data set to find whether the metatdata could
significantly predict the entropy and flux values.
% TODO we haven't explained what ``the 95 performative sessions'' are,
% have we?
In the following analysis, only the 95~``performance context''
sessions were considered, excluding the demonstration and
data-collection contexts. TODO can we justify this?

>>>>

\begin{abstract}
  In this note we describe the analysis of a data archive for an
  ensemble musical performance practice on touch screens where
  transition matrix measures are able to distinguish between sessions
  of different contexts, and the internal structure of these sessions.
  The archive includes 95 touch data logs of ensemble touch-screen
  interactions in concerts, rehearslas and formal HCI studies of
  groups of musicians over more than two years.
  The data analysis has been undertaken by first classifying these
  data into higher level gesture types and summarising the performers'
  transitions between these gestures.
  We find that two transition-matrix measure, flux, and entropy,
  can be used to distinguish different performance types and contexts,
  as well as different phases (beginning, middle and end) of all of
  the performances in our corpus.
\end{abstract}

@@

%\section{General comments}
% \begin{itemize}
% \item we should be clear (and careful) where we're talking about
%   groups TMs, and where we're talking about individual performers
% \item what's the terminology we're using for a ``performance''? Is it
%   ``performance'', or is that too music-specific? ``Session''?
%   Alternatives are ``activity'', or ``interaction'', but they're so
%   general\ldots This decision has implications for our other
%   terminology: participant vs performer vs musician, etc.
% \item Title: maybe ``Traces through time: making sense of longitudinal
%   open-ended interaction data'' or some such? I think that's our big
%   contribution: it takes the in-the-wild, practitioner-driven research
%   program stuff and tries to make some quantitative sense out of it
%   post-hoc.
% \end{itemize}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figures/three-performance-contexts}
  \caption{Collaborative musical interaction with our touch-screen
    interfaces in three different contexts: (left to right) rehearsal
    in a studio,
    performance with percussion instruments, and performance with just
    iPads.\label{fig:three-performance-contexts}}
\end{figure*}

Modern computing devices and interfaces continue to present
opportunities to support new kinds of creative expression and
collaborative interaction~\cite{Resnick:2005yu}. However, it has been
acknowledged that understanding and evaluating interfaces in creative
contexts can be challenging~\cite{Shneiderman:2007qv}. In the context
of live performance, interface evaluation using questionnaires, or
ethnographic data gathering can miss the dynamic need of artists in
the moment of performance and may be impractical in live performative
contexts. When performers use a combuter-based artistic interface,
logged interaction data which has been collected for curation and
preservation~\cite{England:2014ys}, might also allow quantitative
investigation into these performances.

In this note we describe an archive of interaction data logs that has
been accumulated to support a long-term practice-led research program
into simple iPad interfaces for group music-making similar to those
described in Martin et al~\cite{Martin:2014cr}. Over a period of more
than two years (April 2013--June 2015), we collected data from group
music-making sessions where the participants used custom-built,
touch-screen apps on Apple iPad devices. Although several different
apps have been included in our corpus, they shared a common mode of
free-form touch interaction where tapping produced short sounds and
swirling or swiping produced long sounds. Apart from this commonality,
the apps featured a variety of sound palettes, different kinds of
visual feedback, and different kinds of networked interactions
(between the iPads in an ensemble and with a central server).

In all of our interaction sessions, every touch interaction---every
touch down, dragging movement and release---was time-stamped and
logged to disk for later analysis. These sessions included rehearsals
without an audience as well as live performances at arts festivals and
other events. They also included interactions during
formally-structured HCI experiments where participant compared
multiple apps and interfaces. Some of the performances were iPad-only
and others used iPads together with other acoustic percussion
instruments. Although the majority of the sessions were free-form
improvisations, a number of rehearsals and performances involved
composed pieces where the musicians read and performed percussion
notation using the apps.

Representative photographs of some of these session contexts
(rehearsal, performance including other acoustic instruments and iPad
only performance) are are shown
Figure~\ref{fig:three-performance-contexts}) and other descriptive
statistics about the sessions are shown in Table \ref{corpus-table}.
In our corpus, the median length of sessions is 7m29s and the median
number of participants per ensemble is four. Metadata was included
together with the interaction data for each session to record the
\emph{context} (performance, rehearsal, study), session \emph{type}
(improvisation or composition), and \emph{instrumentation}
(iPads-only, or iPads with acoustic percussion). The distribution of
recorded sessions by context and type is shown in Figure
\ref{fig:count-data}.

% It is a feature of creative expression in the arts that highly skilled
% practitioners exploit simple tools to produce works of beauty whether
% it be static paintings or dynamic performance. 
% In previous work~\cite{Martin:2014cr} researchers have shown that
% trained percussionists can exploit a very simple iPad interface to
% discover new categories of musical gestures.
% In that paper, iPad apps
% were studied where tapping was mapped to produce produced short
% sounds, while swiping produced long sounds with the volume
% proportional to the velocity of the moving touch point. 
% It turned out that trained percussionists working together in an
% ensemble were able to discover and repeatedly make use of five macro
% gestures made up of taps and swipes: fast taps, slow taps, random
% taps, short swipes and swirls. This process of discovery of
% interesting musical gestures from a simple interface is analogous to
% what percussionists do with acoustic instruments -- drums and bowls
% and bells are often tapped, stroked and hit in novel ways to
% communicate meaning and emotion through music~\cite{Schick:2006fk}.

% TODO fix this table to remove descriptive data about the
% demonstration sessions.
\begin{table}
\centering
\begin{tabular}{l|llll}
\hline
            & Total & Min  & Median   & Max     \\ 
\hline
N           & 150       &         &        &  \\
%Record size & 306MB     &         &        &   \\
Length      & 20H34M22S & 1M0S    & 7M29S  & 26M0S\\
Participants& 524       & 1       & 4      & 9    \\
Flux        &           & 0.02239 & 0.1445 & 0.3246\\
Entropy     &           &  &  & \\          
\hline
\end{tabular}
\caption{
  Representative data about the corpus of musical touch-screen
  interaction sessions used in this study. These sessions were
  rehearsals, performances, or studies and 
  participants improvised or performed compositions. Each record
  consisted of touch-data recorded in CSV format and was later analysed
  with a gestural classification system. Some participants were
  present in multiple sessions, however this data was not tracked in
  the corpus.\label{corpus-table}}
\end{table}

@@

Suppose that each participant's activity in a collaborative
interaction can at any given time be classified as belonging to one of
$m$ distinct \emph{states}. Over the course of the activity, each
participant $X$ generates a length $N$ sequence of these interaction
states
\begin{equation}
 X_n \hskip 2em n = 1, \ldots, N
\end{equation}
which can be thought of as Markov process. If there is an (automated)
classification process for extracting each participant's state
sequence from (or during) their interaction data, then we can estimate
the $m \times m$ transition matrix $P$ which characterises this
process by simply counting the number of times state $j$ follows state
$i$. The entries of $P$ are then given by:
$p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}$.
% \begin{equation}
%   p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
% \end{equation}
The transition activity of the whole ensemble can be calculated by
averaging the TM for each performer. The matrix can be normalised so
that the sum of all elements is equal to one.

@@

To do this we calculate two high-level quantities from each session:
``flux'' and ``entropy''. The flux of a sequence is defined as the
ratio of state transitions (e.g. $A \rightarrow B$ where $A \neq B$),
to self transitions (e.g. $A \rightarrow A $). The flux of a TM $P$ is
given by: $\mathrm{flux}(P) = 1 - \mathrm{tr}(P)$. Where the trace
$\mathrm{tr}(P)$ is the sum of the diagonal entries of $P$.
Intuitively, flux is a measure of how frequently the participant/group
changes state. Flux returns a value in the interval $[0,1]$ and will
return 0 when participants never change state, and 1 when participants
never stay in the same state for two measurements in a row.

Another useful measure that can be used on a transition matrix is its
entropy, defined in the information-theoretic\cite{Shannon:1948rt}
sense: $H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})$.
% \begin{equation}
%   H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})
% \end{equation}
This entropy measure is small when the matrix is sparse, and largest
when each matrix element is equal. It offers a different perspective
on collaborative interactions than the flux measure by capturing the
breadth of the gestural space explored by the participants throughout
the course of an interaction. Consider the degenerate case of a
participant who alternates between two states for a whole performance:
the flux in this case will be maximal ($\mathrm{flux}(P) = 1$) since
there are no self-transitions, (only $A \rightarrow B$ or
$ B \rightarrow A$) even though the participant has only used a small
subset of the state space. The entropy of this interaction, on the
other hand, will be low. Entropy, therefore, is a measure of how broad
the participant's exploration of the state space is in a given
interaction.

@@

@@

The distribution of flux and entropy with respect to session contexts,
types and instrumentation are shown in Figure
\ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
performed on the data set to find whether the metatdata could
significantly predict the entropy and flux values.
